Synthesizing Quotient Lenses

### 2018-02-05 papers/QuotientLenses/main.tex
Put Quotient Lenses file in ICFP format

\documentclass{svproc}

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}

@@

%%%%%
% Macros
% Colors
\definecolor{dkblue}{rgb}{0,0.1,0.5}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{dkred}{rgb}{0.6,0,0}
\definecolor{dkpurple}{rgb}{0.7,0,0.4}
\definecolor{olive}{rgb}{0.4, 0.4, 0.0}
\definecolor{teal}{rgb}{0.0,0.5,0.5}
\definecolor{orange}{rgb}{0.9,0.6,0.2}
\definecolor{lightyellow}{RGB}{255, 255, 179}
\definecolor{lightgreen}{RGB}{170, 255, 220}
\definecolor{teal}{RGB}{141,211,199}
\definecolor{darkbrown}{RGB}{121,37,0}

@@

% For Inference Rules
\newcommand{\Rule}[2]{\infer{#2}{#1}}
\newcommand{\RuleSide}[3]{\infer[#3]{#2}{#1}}
%\newcommand{\Axiom}[1]{\Rule{}{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\mainmatter              % start of a contribution
%
\title{Synthesizing Quotient Lenses}
%
\titlerunning{Synthesizing Quotient Lenses}  % abbreviated title (for running
% head)                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{}
%
\authorrunning{} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{}
%
\institute{}

\maketitle              % typeset the title of the contribution

\begin{abstract} 
\saz{The old abstract read more like an introduction, so we modified it to be
  the ``new'' introduction.}
\end{abstract}

@@

However, these lens laws are too restrictive for many important transformations.
In particular, these lens laws do not permit transformations that put the data
in canonical forms. In \bibtex{}, an author can be specified as \cd{lastname,
  firstname} and \cd{firstname lastname}. While it is important to allow authors
to be input in both ways, we would like these authors to only be output as
\cd{firstname lastname}, but the lens laws prevent such functions.
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses where
each of the data formats is quotiented by some equivalence relation and the lens
laws are taken modulo equivalence. This both permits canonization functions and
allows lenses to be simpler and more flexible. Inessential details like
whitespace and orderings of fields can be ignored when writing the
transformations between the formats.

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses. Even though quotient lenses permit simpler
transformations between the formats, they merely push complexity elsewhere. To
write a quotient lens, programmers must write with both standard lens
combinators and \emph{canonizers}. Canonizers encode two functions, one that
maps a format to a canonical version of that format, and another that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to enforce that a quotient lens maps
between two different formats, they must write regular expressions specifying
such formats. In this way, writing quotient lenses involves writing 3 terms: (1)
the data formats on each side of the lens, (2) canonizers to normalize the data
and choose representatives, and (3) a lens between the data formats when in
canonical form, even though, typically, these three share a great deal of
structure. Moreover, writing the lens between the equivalence classes can be
challenging due to the rather fiddly typechecking, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression while
simultaneously specifying the canonical representatives on the regular
expression's language. With QREs, programmers needn't define the data and
canonization functions separately -- by adding simple annotations to the data,
the canonization functions can be inferred.

Secondly, we use QREs to define a class of quotient lenses called {\em QRE
  lenses} which map betweeen QREs, and show how to synthesize QRE lenses. Using
this tool, a lens programmer need not manually derive the lens herself, she
merely must provide a pair of QREs (describing the formats the lens should map
between), and a suite of examples (demonstrating how the lens should act on
specific instances).  The resulting lens is guaranteed to map between the
languages of the QREs and satisfy the provided examples.

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions} (QREs)
  that provide a compact, convenient notation for regular languages and their
  canonical representatives (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs
and QRE lens synthesis.  We demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\bibliographystyle{plain}
\bibliography{local,bcp}

>>>>

\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For double-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
% % For single-blind review submission, w/o CCS and ACM Reference (max
% submission space)
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For single-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true} % For
% final camera-ready submission, w/ required CCS and ACM Reference
% \documentclass[acmsmall]{acmart}\settopmatter{}

% % Journal information % Supplied to authors by publisher for camera-ready
% submission; % use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{ICDP} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}

@@

% %%%% Macros Colors
\definecolor{dkblue}{rgb}{0,0.1,0.5}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{dkred}{rgb}{0.6,0,0}
\definecolor{dkpurple}{rgb}{0.7,0,0.4}
\definecolor{olive}{rgb}{0.4, 0.4, 0.0}
\definecolor{teal}{rgb}{0.0,0.5,0.5}
\definecolor{orange}{rgb}{0.9,0.6,0.2}
\definecolor{lightyellow}{RGB}{255, 255, 179}
\definecolor{lightgreen}{RGB}{170, 255, 220}
\definecolor{teal}{RGB}{141,211,199}
\definecolor{darkbrown}{RGB}{121,37,0}

@@

% For Inference Rules
\newcommand{\Rule}[2]{\infer{#2}{#1}}
\newcommand{\RuleSide}[3]{\infer[#3]{#2}{#1}}
% \newcommand{\Axiom}[1]{\Rule{}{#1}}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
\saz{The old abstract read more like an introduction, so we modified it to be
the ``new'' introduction.}
\end{abstract}

\keywords{quotient lenses, synthesis}
\maketitle              % typeset the title of the contribution

@@

However, these lens laws are too restrictive for many important transformations.
In particular, these lens laws do not permit transformations that put the data
in canonical forms. In \bibtex{}, an author can be specified as \cd{lastname,
firstname} and \cd{firstname lastname}. While it is important to allow authors
to be input in both ways, we would like these authors to only be output as
\cd{firstname lastname}, but the lens laws prevent such functions.
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses where
each of the data formats is quotiented by some equivalence relation and the lens
laws are taken modulo equivalence. This allows lenses to be simpler and more
flexible. For example, inessential details like whitespace and orderings of
fields can be ignored when writing the transformations between the formats.

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses. Even though quotient lenses permit simpler
transformations between the formats, they merely push complexity elsewhere. To
write a quotient lens, programmers must write with standard lens
combinators and \emph{canonizers}. Canonizers encode two functions, one that
maps a format to a canonical version of that format, and another that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to enforce that a quotient lens maps
between two different formats, they must write regular expressions specifying
such formats. In this way, writing quotient lenses involves writing 3 terms: (1)
the data formats on each side of the lens, (2) canonizers to normalize the data
and choose representatives, and (3) a lens between the data formats when in
canonical form, even though, typically, these three share a great deal of
structure. Moreover, writing the lens between the equivalence classes can be
challenging due to the rather fiddly typechecking, mainly due to
``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives on the regular expression's language. With
QREs, programmers do not need to define the data and canonization functions
separately -- by adding simple annotations to the data, the canonization
functions can be inferred.

Second, we use QREs to define a class of quotient lenses called {\em QRE
lenses} which map betweeen QREs, and show how to synthesize QRE lenses. In
order to derive a lens, a programmer need not manually derive the lens
herself, she only needs to provide a pair of QREs describing the formats the
lens should map between, and a suite of examples that demonstrates how the
lens should act on specific examples.  The resulting lens is guaranteed to
map between the languages of the QREs and satisfy the provided examples.

In summary, our main contributions are:
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs)
  which provide a compact, convenient notation that expresses an equivalence
  relation on a regular languages as well as as a canonical set representatives
  for this equavalence relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}, a subset of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution in this
  paper is a proof that each QRE lens $q$ defines the same transformation as a
  QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
  \item Using this normal form, we reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate its applicability by using it to
  synthesize QRE lenses between several real-world data formats from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\bibliographystyle{plain}
\bibliography{local}
